---
title: "Data 607 Final Project Proposal"
author: "Zaneta Paulusova & Inna Yedzinovich"
output: powerpoint_presentation

---

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(usmap)
```

### Introduction: 
In this study, we will explore the relationship between the number of drivers and the incidence of fatal accidents in the USA. Our analysis will focus on fatality data from the National Highway Traffic Safety Administration and driving population statistics from the Federal Highway Administration for the years 2018, 2020, and 2022. Specifically, we will compare data from the pre-COVID year (2018), the COVID-19 year (2020), and the post-COVID year (2022). We aim to determine whether an increase in the driving population correlates with a rise in the number of fatalities. Our hypothesis is that a larger number of drivers will lead to a higher incidence of fatal accidents.

### Data Source 
Data of fatal accidents per state 

 - https://www-fars.nhtsa.dot.gov/states/statesfatalitiesfatalityrates.aspx

Population of drivers per state:

 - 2022 - https://www.fhwa.dot.gov/policyinformation/statistics/2022/dl1c.cfm
 
 - 2020 - https://www.fhwa.dot.gov/policyinformation/statistics/2020/dl1c.cfm
 
 - 2018 - https://www.fhwa.dot.gov/policyinformation/statistics/2018/dl1c.cfm

### Fatality Data

```{r}
Fatality_data <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Fatalities%20and%20Fatality%20Rates%20by%20STATE%2C%201994%20-%202022%20-%20State%20USA.csv")
Fatality_data 
```

```{r}
Fatalitydata <- Fatality_data[, c("State", "X2018", "X2020", "X2022")]
names(Fatalitydata) <- c("State", "Y18", "Y20", "Y22")
Fatalitydata <- Fatalitydata[-c(52), ]
head(Fatalitydata)
```

```{r}
 Fatalitydata$State[Fatalitydata$State=="District of Columbia"] <-"Dist. of Col."
Fatalitydata
```


### Driving Population for Year 2022

```{r}
data_2022 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202022.csv")
data_2022
```

``` {r}
names(data_2022) <- c("State", "Male_Drivers_Y22", "Male_Driver_%_Y22", "Female_Driver_Y22", "Female_Driver_%_Y22", "Total_Drivers_Y22", "Commercial_Motor_Vehicles_Registered_Y22", "Population_Resident_Y22", "Population_Male_Y22", "Population_Female_Y22", "Population_Total_Y22", "Drivers_Per_1K_Total_Resident_Polulation_Y22", "Drivers_Per_1K_Age_Population_Y22")
data_2022
data22 <- data_2022[-c(1,2,3,4,5,57,58,59,60,61,62,63,64), ]
data22
```

Remove fotenote from State

```{r}
data22$State[data22$State=="Alaska(2)"] <-"Alaska"
data22$State[data22$State=="Arkansas(4)"] <-"Arkansas"
data22$State[data22$State=="Connecticut(7)"] <-"Connecticut"
data22$State[data22$State=="Dist. of Col.(4)"] <-"Dist. of Col."
data22$State[data22$State=="Colorado(6)"] <-"Colorado"
data22$State[data22$State=="Hawaii(7)"] <-"Hawaii"
data22$State[data22$State=="Iowa(5)"] <-"Iowa"
data22$State[data22$State=="Maryland(6)"] <-"Maryland"
data22$State[data22$State=="Michigan(6)"] <-"Michigan"
data22$State[data22$State=="Minnesota(6)"] <-"Minnesota"
data22$State[data22$State=="Nevada(6)"] <-"Nevada"
data22$State[data22$State=="New Hampshire(4)(6)"] <-"New Hampshire"
data22$State[data22$State=="New Jersey(6)"] <-"New Jersey"
data22$State[data22$State=="New Mexico(6)"] <-"New Mexico"
data22$State[data22$State=="Oregon(6)"] <-"Oregon"
data22$State[data22$State=="Vermont(6)"] <-"Vermont"
data22$State[data22$State=="Washington(6)"] <-"Washington"
data22$State[data22$State=="Wyoming(5)"] <-"Wyoming"
head(data22)
```


### Driving Population for Year 2020

``` {r}
data_2020 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202020.csv")
head(data_2020)
```

```{r}
names(data_2020) <- c("State", "Male_Drivers_Y20", "Male_Driver_%_Y20", "Female_Driver_Y20", "Female_Driver_%_Y20", "Total_Drivers_Y20", "Commercial_Motor_Vehicles_Registered_Y20", "Population_Resident_Y20", "Population_Male_Y20", "Population_Female_Y20", "Population_Total_Y20", "Drivers_Per_1K_Total_Resident_Polulation_Y20", "Drivers_Per_1K_Age_Population_Y20")
data20 <- data_2020[-c(1,2,3,4,5,6,58,59,60,61,62,63,64), ]
head(data20)
```

```{r}
data20$State[data20$State=="Arkansas(5)"] <-"Arkansas"
data20$State[data20$State=="New York(5)"] <-"New York"
head(data20)
```


### Driving Population for Year 2018

``` {r}
data_2018 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202018.csv")
head(data_2018)
```

```{r}
names(data_2018) <- c("State", "Male_Drivers_Y18", "Male_Driver_%_Y18", "Female_Driver_Y18", "Female_Driver_%_Y18", "Total_Drivers_Y18", "Commercial_Motor_Vehicles_Registered_Y18", "Population_Resident_Y18", "Population_Male_Y18", "Population_Female_Y18", "Population_Total_Y18", "Drivers_Per_1K_Total_Resident_Polulation_Y18", "Drivers_Per_1K_Age_Population_Y18")
data18 <- data_2018[-c(1,2,3,4,5,6,58,59,60), ]
data18
```

```{r}
data18$State[data18$State=="Alaska2/"] <-"Alaska"
data18$State[data18$State=="Illinois2/"] <-"Illinois"
data18
```


### Merging and Cleaning Data

``` {r}
df22 <- data22[, c("State", "Total_Drivers_Y22")]
df20 <- data20[, c("State", "Total_Drivers_Y20")]
df18 <- data18[, c("State", "Total_Drivers_Y18")]

df_pop <- Reduce(merge,list(df22,df20,df18))
df_pop

```

```{r}
dp <- df_pop %>%
  pivot_longer(cols = c('Total_Drivers_Y22', 'Total_Drivers_Y20', 'Total_Drivers_Y18'), names_to = "Total_Drivers_Per_Years", values_to = "Drivers_Population")
dp
```


### Descriptive Statistics

``` {r}
summary(dp)
```

### Graph of Fatalities

``` {r}
fd <- Fatalitydata %>%
  pivot_longer(cols = c('Y18', 'Y20', 'Y22'), names_to = "Years", values_to = "Fatalities_Rate")

ggplot((data = fd), aes(x = State, y = Fatalities_Rate, fill = Years)) + 
  geom_col(position = position_dodge()) + 
  ggtitle("Fatalities Rate for Years 2018, 2020, & 2022") +
  theme(axis.text.x = element_text(angle = 66, hjust = 1))
```

This graph provides a clear visual representation of fatalities rates across states and years, allowing for comparisons and analysis of trends.

The graph effectively visualizes traffic fatalities across US states for 2018, 2020, and 2022, highlighting significant variations between populous states like California, Texas, and Florida compared to smaller states.

Key Observations:
 - High Fatalities States: California, Florida, and Texas consistently have the highest fatalities rates.
 - Low Fatalities States: Vermont, Wyoming, and Rhode Island have the lowest rates.
 - Yearly Changes: While some states show slight increases or decreases, the overall trend varies by state.

### Graph of Drivers Population

```{r}
dpop <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/dpop.csv")
head(dpop)
```

```{r}
dp <- df_pop %>%
  pivot_longer(cols = c('Total_Drivers_Y22', 'Total_Drivers_Y20', 'Total_Drivers_Y18'), names_to = "Total_Drivers_Per_Years", values_to = "Drivers_Population") %>%
  mutate(Total_Drivers_Per_Years = recode(Total_Drivers_Per_Years,
                                          "Total_Drivers_Y22" = "2022",
                                          "Total_Drivers_Y20" = "2020",
                                          "Total_Drivers_Y18" = "2018"),
         Drivers_Population = as.numeric(gsub("[^0-9]", "", Drivers_Population)))

ggplot(data = dp, aes(x = State, y = Drivers_Population, fill = Total_Drivers_Per_Years)) + 
  geom_col(position = position_dodge()) + 
  ggtitle("Graph of Drivers Population") +
  theme(axis.text.x = element_text(angle = 66, hjust = 1)) +
  labs(y = "Drivers Population")
```

This is a bar graph showing "Drivers Population for the Years 2018, 2020, & 2022" across different U.S. states. Here are the key observations:

Key Observations:
 - The largest populations are around 27 million drivers
 - The smallest populations appear to be less than 1 million drivers
 - Most states fall in the range of 2-8 million drivers
 
 
## Conclusion 

Let's see which year has the most fatalities: 

``` {r}
total_population_by_year <- dp %>%
  group_by(Total_Drivers_Per_Years) %>%
  summarise(Total_Drivers = sum(Drivers_Population, na.rm = TRUE))
total_population_by_year

year_with_most_fatalities <- total_population_by_year %>%
  filter(Total_Drivers == max(Total_Drivers)) %>%
  pull(Total_Drivers_Per_Years)

year_with_least_fatalities <- total_population_by_year %>%
  filter(Total_Drivers == min(Total_Drivers)) %>%
  pull(Total_Drivers_Per_Years)

print(paste("The year with the most fatalities is", year_with_most_fatalities))
print(paste("The year with the least fatalities is", year_with_least_fatalities))
```

State the null hypothesis and perform t-tests:
Null hypothesis: COVID time 2020 has fewer fatalities than after COVID 2022 and pre-COVID 2018.

``` {r}
null_hypothesis <- "COVID time 2020 has fewer fatalities than after COVID 2022 and pre-COVID 2018."

fatalities_2020 <- dp %>% filter(Total_Drivers_Per_Years == "2020") %>% pull(Drivers_Population)
fatalities_2018 <- dp %>% filter(Total_Drivers_Per_Years == "2018") %>% pull(Drivers_Population)
fatalities_2022 <- dp %>% filter(Total_Drivers_Per_Years == "2022") %>% pull(Drivers_Population)

t_test_2018_2020 <- t.test(fatalities_2018, fatalities_2020)
t_test_2020_2022 <- t.test(fatalities_2020, fatalities_2022)

print(t_test_2018_2020)
print(t_test_2020_2022)

if (t_test_2018_2020$p.value < 0.05 & t_test_2020_2022$p.value < 0.05) {
  print("The null hypothesis is disproved. COVID time 2020 does not have fewer fatalities than after COVID 2022 and pre-COVID 2018.")
} else {
  print("The null hypothesis is not disproved. COVID time 2020 has fewer fatalities than after COVID 2022 and pre-COVID 2018.")
}
```

 
Based on the Welch Two Sample t-tests, there is no significant difference in the mean number of fatalities between the years 2018 and 2020 (p-value = 1) or between 2020 and 2022 (p-value = 0.9). The confidence intervals for both tests include zero, indicating that the differences in means are not statistically significant.

Therefore, the null hypothesis is not disproved: the COVID-19 year (2020) does not have significantly fewer fatalities compared to the pre-COVID year (2018) and the post-COVID year (2022). 
 
