---
title: "Data 607 Final Project Proposal"
author: "Zaneta Paulusova & Inna Yedzinovich"
output: pdf_document
---

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(usmap)
```


We will examine the relationship between the number of drivers and the incidence of fatal accidents in the USA. Our focus will be on fatality data from the National Highway Traffic Safety Administration and driving population statistics from the Federal Highway Administration for the years 2018, 2020, and 2022. Our analysis will investigate whether the number of fatalities increases as the driving population grows. I believe that an increase in the number of drivers is likely to lead to an increase in the number of fatalities as well.

Data Source:
data of fatal accidents per state – https://www-fars.nhtsa.dot.gov/states/statesfatalitiesfatalityrates.aspx

population of drivers per state:
2022 - https://www.fhwa.dot.gov/policyinformation/statistics/2022/dl1c.cfm
2020 - https://www.fhwa.dot.gov/policyinformation/statistics/2020/dl1c.cfm
2018 - https://www.fhwa.dot.gov/policyinformation/statistics/2018/dl1c.cfm


#Data Source 

###Fatality Data
```{r}
Fatality_data <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Fatalities%20and%20Fatality%20Rates%20by%20STATE%2C%201994%20-%202022%20-%20State%20USA.csv")
head(Fatality_data)
```
```{r}
colnames(Fatality_data)
```

###Diving Population for Year 2022
```{r}
data_2022 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202022.csv")
head(data_2022)
```


```{r}
colnames(data_2022)
```

###Adding a description for each column.
```{r}
d_22<-data_2022
names(d_22)[1]="State"
names(d_22)[2]="Male_Drivers_Y22"
names(d_22)[3]="Male_Driver_%_Y22"
names(d_22)[4]="Female_Driver_Y22"
names(d_22)[5]="Female_Driver_%_Y22"
names(d_22)[6]="Total_Drivers_Y22"
names(d_22)[7]="Commercial_Motor_Vehicles_Registered_Y22"
names(d_22)[8]="Population_Resident_Y22"
names(d_22)[9]="Population_Male_Y22"
names(d_22)[10]="Population_Female_Y22"
names(d_22)[11]="Population_Total_Y22"
names(d_22)[12]="Drivers_Per_1K_Total_Resident_Polulation_Y22"
names(d_22)[13]="Drivers_Per_1K_Age_Population_Y22"
colnames(d_22)
```

```{r}
d_22
```

###Taking out blank and not data related information rows
```{r}
d22<-d_22[-c(1,2,3,4,5,57,58,59,60,61,62,63,64), ] 
d22
```

###Diving Population for Year 2020
```{r}
data_2020 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202020.csv")
head(data_2020)
```

```{r}
colnames(data_2020)
```

###Adding a description for each column.
```{r}
d_20<-data_2020
names(d_20)[1]="State"
names(d_20)[2]="Male_Drivers_Y20"
names(d_20)[3]="Male_Driver_%_Y20"
names(d_20)[4]="Female_Driver_Y20"
names(d_20)[5]="Female_Driver_%_Y20"
names(d_20)[6]="Total_Drivers_Y20"
names(d_20)[7]="Commercial_Motor_Vehicles_Registered_Y20"
names(d_20)[8]="Population_Resident_Y20"
names(d_20)[9]="Population_Male_Y20"
names(d_20)[10]="Population_Female_Y20"
names(d_20)[11]="Population_Total_Y20"
names(d_20)[12]="Drivers_Per_1K_Total_Resident_Polulation_Y20"
names(d_20)[13]="Drivers_Per_1K_Age_Population_Y20"
colnames(d_20)
```
```{r}
d_20
```

###Taking out blank and not data related information rows
```{r}
d20<-d_20[-c(1,2,3,4,5,6,58,59,60,61,62,63,64), ] 
d20
```

###Diving Population for Year 2018
```{r}
data_2018 <- read.csv("https://raw.githubusercontent.com/ZanetaP02/DAta607-Final-Project/refs/heads/main/Licensed%20Drivers%20by%20Sex%20and%20Ratio%20to%20Population%20-%202018.csv")
head(data_2018)
```
```{r}
colnames(data_2018)
```
###Adding a description for each column.
```{r}
d_18<-data_2018
names(d_18)[1]="State"
names(d_18)[2]="Male_Drivers_Y18"
names(d_18)[3]="Male_Driver_%_Y18"
names(d_18)[4]="Female_Driver_Y18"
names(d_18)[5]="Female_Driver_%_Y18"
names(d_18)[6]="Total_Drivers_Y18"
names(d_18)[7]="Commercial_Motor_Vehicles_Registered_Y18"
names(d_18)[8]="Population_Resident_Y18"
names(d_18)[9]="Population_Male_Y18"
names(d_18)[10]="Population_Female_Y18"
names(d_18)[11]="Population_Total_Y18"
names(d_18)[12]="Drivers_Per_1K_Total_Resident_Polulation_Y18"
names(d_18)[13]="Drivers_Per_1K_Age_Population_Y18"
colnames(d_18)
```
```{r}
d_18
```

###Taking out blank and not data related information rows
```{r}
d18<-d_18[-c(1,2,3,4,5,6,58,59,60), ] 
d18
```


##Edit Data Source

We are only looking for fatality data from the years 2018, 2020, and 2022, so we need a table that includes information for those years and the states.
```{r}
fatal_data <-Fatality_data[,c("State","X2018","X2020", "X2022")]
fatal_data
```

###Updating a description for column.
```{r}
f_d<-fatal_data
names(f_d)[1]="State"
names(f_d)[2]="Y18"
names(f_d)[3]="Y20"
names(f_d)[4]="Y22"
f_d
```

```{r}
colnames(f_d)
```
###Taking out blank and not data related information rows
```{r}
f_d1<-f_d[-c(52), ] 
f_d1
```
###Editing only particle colums in thw 2018, 2020, and 2022 data
```{r}
df22<-d22[,c("State","Male_Drivers_Y22","Female_Driver_Y22", "Total_Drivers_Y22","Population_Resident_Y22")]
df22
```



rownames(df22)
```{r}
df20 <-d20[,c("State","Male_Drivers_Y20","Female_Driver_Y20", "Total_Drivers_Y20","Population_Resident_Y20")]
df20
```
```{r}
df18 <-d18[,c("State","Male_Drivers_Y18","Female_Driver_Y18", "Total_Drivers_Y18","Population_Resident_Y18")]
df18
```




#Graph of Fatalities

###Moving the years in to one column
```{r}
fd <- f_d1%>%
  pivot_longer(cols = c('Y18', 'Y20', 'Y22'), names_to = "Years", values_to = "Fatalities_Rate")
fd
```

```{r}
ggplot((data = fd), aes(x = State, y = Fatalities_Rate, fill = Years)) + geom_col(position = position_dodge())+ggtitle("Fetalities Rate for Years 2018, 2020, & 2022") +
theme(axis.text.x = element_text(angle = 66, hjust = 1))
```

```{r}
pop_d <-merge(df22, df20,by="State", all = TRUE)
pop_d
```

```{r}
pop_drive <-merge(pop_d, df18,by="State", all = TRUE)
pop_drive
```


```{r}
pop_driver <- pop_drive%>%
  pivot_longer(cols = c('Total_Drivers_Y22', 'Total_Drivers_Y20', 'Total_Drivers_Y18'), names_to = "Total_Drivers_Per_Years", values_to = "Drivers_Population")
pop_driver
```

```{r}
drivers_pop <-pop_driver[,c("State","Total_Drivers_Per_Years","Drivers_Population")]
drivers_pop
```


```{r}
dp <- na.omit(drivers_pop)
dp
```


```{r}
ggplot((data = dp), aes(x = State, y = Drivers_Population, fill = Total_Drivers_Per_Years)) + geom_col(position = position_dodge())+ggtitle("Drivers Population for the Years 2018, 2020, & 2022") +
theme(axis.text.x = element_text(angle = 66, hjust = 1))
```

